A general Nullstellensatz for generalized spaces

### 2019-05-03 paper-qcoh.tex
Slides almost finished
to do
More work on the introduction

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\begin{abstract}
  We give a general Nullstellensatz for the generic model of a
  geometric theory, useful as a source of nongeometric sequents validated by
  the generic model, and characterize the first-order and higher-order formulas validated by the
  genetric model.
\end{abstract}

@@

@@

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, of local rings or of intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms, consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide:
\begin{enumerate}
\item The sequent~$\sigma$ is provable modulo~$\TT$.
\item The sequent~$\sigma$ holds for any~$\TT$-model in any Grothendieck topos.
\item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate}
One could argue that it is this model which we
implicitly refer to when we utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

@@

One of the most celebrated such sequents arises in the case that~$\TT$ is the
theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
the topos of sheaves over the site of schemes locally of finite presentation,
and the generic model is the functor~$\affl$ of points of the affine line, the
functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom(X,\mathbb{A}^1) =
\O_X(X)$.

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
any nonzero element is invertible.
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings which is expressible as a geometric sequent,
suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

A related nongeometric sequent is valid in the little Zariski topos of the
spectrum of a ring~$A$, the classifying topos of local localizations of~$A$.
If~$A$ is reduced, the generic model validates the dual condition that any
noninvertible element is zero. This property has been used to give a short and
even constructive proof of Grothendieck's generic freeness lemma,
substantially improving on previously published
proofs~\cite{blechschmidt:generic-freeness}.

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero. \smallskip
\item The Nullstellensatz holds:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
\item $\affl$ is \emph{synthetically quasicoherent}: For any
finitely presentable~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
of~$\affl$-algebra homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
% Take care: In sect:applications, we state that each of these nongeometric
% sequents also holds in the ring classifier.
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language of the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

\paragraph{Characterizing nongeometric sequents} Referring to one of the previous
examples, Tierney remarked around the time that those sequents were first
studied that ``[it] is surely important, though its precise significance is
still somewhat obscure---as is the case with many such nongeometric
formulas''~\cite[p.~209]{tierney:spectrum}. In view of their importance, is
there a way to discover nongeometric sequents in a systematic fashion? To
characterize the nongeometric sequents holding in classifying toposes? To this
end, Wraith put forward a specific conjecture~\cite[p.~336]{wraith:intuitionistic-algebra}:
\begin{quote}
\emph{The problem of characterising all the non-geometric properties of a generic
model appears to be difficult. If the generic model of a geometric theory~$\TT$
satisfies a sentence~$\alpha$ then any geometric consequence of $\TT + \alpha$ has to be a
consequence of~$\TT$. We might call~$\alpha$ $\TT$-redundant. Does the
generic~$\TT$-model satisfy all~$\TT$-redundant sentences?}
\end{quote}
This question was recently answered in the negative by Bezem, Buchholtz and
Coquand~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}; hence the
characterization we propose is necessarily more nuanced.

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory. We then have, subject to some qualifications made precise in
Section~\ref{sect:main}, the following general Nullstellensatz:

\begin{thm}\label{thm:nullstellensatz0}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{equation}\tag{$\dagger$}\label{eq:nullstellensatz0}
\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
\end{equation}
\end{thm}

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \? U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz0} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
substantially broadens the scope of the Nullstellensatz.

@@

@@

@@

\paragraph{Acknowledgments} XXX

@@

Johnstone~\cite{XXX} lists several nongeometric sequents(XXX) holding in the
classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
no function symbols, relations symbols or axioms, including:
\begin{align*}
  \forall x,y\?U_\OO\_ \neg\neg(x = y) &&
  \forall x_1,\ldots,x_n\?U_\OO\_ \neg \forall y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)
\end{align*}
Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
the second expresses that~$U_\OO$ is close to being infinite. Using the
Nullstellensatz, these formulas can be verified as follows.

@@

* XXX Discuss related work, especially: Olivia 6.1.3, ...

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx,color}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\begin{abstract}
  We give a general Nullstellensatz for the generic model of a
  geometric theory, useful as a source of nongeometric sequents validated by
  the generic model, and characterize the first-order and higher-order formulas validated by the
  generic model.
\end{abstract}

@@

\begin{center}
  \textcolor{red}{\textbf{-- rough draft --}}

  \emph{comments welcome at ingo.blechschmidt@univr.it}
\end{center}

@@

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, the theory of local rings or the theory of intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
formula} we will mean a formula which may contain, additionally to the
connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
and~${\forall}$.

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$.
This model is \emph{conservative} in that
for any geometric sequent~$\sigma$, the following notions
coincide:
\begin{enumerate}
\item The sequent~$\sigma$ is provable modulo~$\TT$.
\item The sequent~$\sigma$ holds for any~$\TT$-model in any Grothendieck topos.
\item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate}
One could argue that it is this model which we
implicitly refer to when we utter the phrase ``Let~$M$ be
a~$\TT$-model.''.\footnote{For instance, this point of view is fundamental to
the slogan \emph{continuity is geometricity}, as stressed by
Vickers~\cite{vickers:continuity}.}
It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

@@

One of the most celebrated such sequents arises in the case that~$\TT$ is the
theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
the topos of sheaves over the site of schemes locally of finite presentation,
and the generic model is the functor~$\affl$ of points of the affine line, the
functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
\O_X(X)$.

From the point of view of the Zariski topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
any nonzero element is invertible.
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings which is expressible as a geometric sequent,
suppose without loss of generality that the field axioms holds.
This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

We surmise that many more reduction techniques along these lines exist for
other kinds of algebraic objects. However, when actually using such techniques
in practice, we face the challenge that while we can use them to prove results
about \emph{all} local rings, \emph{all} modules and so on, it is difficult to
incorporate specific information about a particular local ring or a particular
module at hand. This difficulty is compounded by the fact that interesting
nongeometric properties are typically not inherited by the generic model of
quotient theories -- for instance the generic ring validates the formula~$\forall
x\?U_\TT\_ \neg\neg(x = 0)$ while the generic local ring does not.

Hence it is useful to turn to geometric theories which refer to a given
mathematical object. For instance, given a ring~$A$, there is the theory of
local localizations of~$A$, and its classifying topos is known in algebraic
geometry as the little Zariski topos of~$A$, the topos of sheaves over the
spectrum of~$A$. If~$A$ is reduced, this topos validates the dual condition that any
noninvertible element is zero. This property has been used to give a short and
even constructive proof of Grothendieck's generic freeness lemma,
substantially improving on previously published
proofs~\cite{blechschmidt:generic-freeness}.

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero. \smallskip
\item The Nullstellensatz holds:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
\item $\affl$ is \emph{synthetically quasicoherent}: For any
finitely presentable~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
of~$\affl$-algebra homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
% Take care: In sect:applications, we state that each of these nongeometric
% sequents also holds in the ring classifier.
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language of the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.

Nontrivial nongeometric sequents are not an exclusive feature of sheaf toposes.
The generic model of a theory of presheaf type typically also validates such
sequents. For instance, all of the aforementioned properties of the generic
local ring are shared by the generic ring, which lives in the presheaf
topos~$[\mathrm{Ring}_\mathrm{fp}, \Set]$.
\medskip

\paragraph{Characterizing nongeometric sequents} Referring to the field condition in the little Zariski topos,
Tierney remarked around the time that those sequents were first
studied that ``[it] is surely important, though its precise significance is
still somewhat obscure---as is the case with many such nongeometric
formulas''~\cite[p.~209]{tierney:spectrum}. In view of their importance, is
there a way to discover nongeometric sequents in a systematic fashion? To
characterize the nongeometric sequents holding in classifying toposes? To this
end, Wraith put forward a specific conjecture~\cite[p.~336]{wraith:intuitionistic-algebra}:
\begin{quote}
\emph{The problem of characterising all the non-geometric properties of a generic
model appears to be difficult. If the generic model of a geometric theory~$\TT$
satisfies a sentence~$\alpha$ then any geometric consequence of $\TT + \alpha$ has to be a
consequence of~$\TT$. We might call~$\alpha$ $\TT$-redundant. Does the
generic~$\TT$-model satisfy all~$\TT$-redundant sentences?}
\end{quote}
Because classical logic is conservative over intuitionistic logic for geometric
sequents, this question has a trivial negative answer: No, any instance of the
law of excluded middle over the signature of~$\TT$ is~$\TT$-redundant but
typically not validated by the generic~$\TT$-model. Moreover,
Bezem, Buchholtz and
Coquand recently showed that the answer is still negative even if appropriate
care is taken to exclude these
counterexamples~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}. Hence our
characterization of the first-order theory validated by the generic~$\TT$-model
is necessarily more nuanced.

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ varies
with the slice, as~$U_\TT$ changes form on different stages. This mismatch is solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory. We then have, subject to some qualifications made precise in
Section~\ref{sect:main}, the following general Nullstellensatz:

\begin{thm}\label{thm:nullstellensatz0}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{equation}\tag{$\dagger$}\label{eq:nullstellensatz0}
\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
if~$\ul{\TT}/U_\TT$ proves~$\sigma$.}
\end{equation}
\end{thm}

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$ with coefficients in~$U_\TT$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \? U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz0} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
broadens the scope of the Nullstellensatz and is, whenever applicable, useful
for simplifying computations.

@@

It is not entirely obvious how to correctly phrase
axiom~\eqref{eq:nullstellensatz0} in first-order logic, an issue
which we will discuss when we restate Theorem~\ref{thm:characterization0} in
the body of this note (Theorem~\ref{thm:characterization}).

@@

% XXX: mention internal language in general, and SDG?
\paragraph{Related work} The topos-theoretic Nullstellensatz is related to
several precursors. A corollary of the Nullstellensatz is that, over the
first-order theory validated by~$U_\TT$, any first-order formula is in fact
logically equivalent to a geometric formula. This corollary has already been
observed by Butz and Johnstone~\cite[Lemma~4.2]{butz-johnstone:first-order}. At
that point, a characterisation of the first-order formulas in the general case,
of the form as in Theorem~\ref{thm:characterization0}, was still missing.

The higher-order variant of our Nullstellensatz
(Theorem~\ref{thm:definability}) relativizes Cara\-mel\-lo's completeness
theorem~\cite[Theorem~2.4(ii)]{caramello:definability}. Her theorem is the
external statement that any subobject of~$U_\TT$ is given by the interpretation
of a geometric formula over the signature of~$\TT$; our relativization states
that, internally to~$\Set[\TT]$, any subset of~$U_\TT$ is given by a
geometric$^\star$ formula over the signature of~$\ul{\TT}/U_\TT$. As in the
first-order case, the passage from the external to the internal phrasing
necessitates the switch from~$\TT$ to~$\ul{\TT}/U_\TT$.

The Nullstellensatz yields for a given (perhaps conditional) truth about the
generic model a proof modulo the axioms of~$\ul{\TT}/U_\TT$. As illustrated in
Section~\ref{sect:applications}, the procedure we typically use to extract
fruitful information from such a proof is to apply it to a
further~$\ul{\TT}/U_\TT$-model which we specifically construct for the purposes
at hand. Hence we go from truth in~$U_\TT$ via provability to truth in a
further model, that is, we use provability as a (one-way) \emph{bridge}:
\[ \xymatrix{
  & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
  \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
  \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
} \]
This perspective is inspired by Caramello's research
program~\cite{caramello:tst}, though it is not an instance of her main technical
device for establishing bridges, namely exploiting the fact that a single topos
may admit descriptions using quite different sites.
\medskip

\paragraph{Outlook} The Nullstellensatz yields a description of the first-order
and higher-order theory validated by the generic model of a geometric theory.
As already indicated, with more examples and details to be given in
Section~\ref{sect:applications}, this description explains and puts into
perspective several established results and observations.

From the point of view of applications, the main task for the future is to
explore the description in the case of particular geometric theories of
interest. Just as the reduction technique ``any reduced ring is a field'' --
valid because the generic localization of a reduced ring is a field -- enabled
a short and simple proof of Grothendieck's generic freeness
lemma~\cite[Section~11.5]{blechschmidt:phd}, more reduction techniques along
these lines should exist in a wide range of subjects. We hope that with the
Nullstellensatz at hand, discovering such techniques can progress in a more
systematic fashion.

From the point of view of theory, the appropriate context for the
Nullstellensatz should be determined. We formulate the Nullstellensatz in the
context of geometric theories and their classifying Grothendieck toposes, but
there should be analogues for the Nullstellensatz in other contexts where it
makes sense to speak of ``classifying gadgets'', and these analogues should
shed light on the topos-theoretic version and ideally even explain it from deeper
principles. In particular, preliminary computations indicate that there is a
version of the Nullstellensatz for arithmetic universes, the predicative cousin
of toposes introduced by Joyal~\cite{XXX} and recently an important object of
consideration by Maietti and Vickers~\cite{XXX}.
\medskip

@@

\paragraph{Acknowledgments} XXX
% don't forget: Thierry, and Steve and Martín for 6WFTop; also of course
% Alexander. Also include grant statement.

@@

Butz and Johnstone~\cite{butz-johnstone:first-order} list several nongeometric sequents holding in the
classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
no function symbols, relations symbols or axioms, including:
\begin{align*}
  \forall x,y\?U_\OO\_ \neg\neg(x = y) &&
  \forall x_1,\ldots,x_n\?U_\OO\_ \neg \forall y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)
\end{align*}
Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
the second expresses that~$U_\OO$ is close to being infinite. Using the
Nullstellensatz, these formulas can be verified as follows.

@@

@@

* XXX Upgrade Horn section to cartesian theories?
